# Cell-cell communication {#sec-img-cell-cell-communication}

## Preamble

### Introduction

Cell-cell communication analyses aim to investigate how cells interact within tissues through signaling pathways and communication networks.

Several methods have been developed for these types of analyses. In general, these methods require annotated cell types, spatial coordinates, and curated lists of ligand-receptor (L-R) pairs (and optionaly pathway or target gene sets) as inputs. Ligand-receptor pairs can be accessed from databases including `OmniPath`, `CellChatDB`, `CellPhoneDB`, and others.


### Dependencies

```{r deps, message=FALSE, warning=FALSE}
library(reticulate)
library(zellkonverter)
library(SingleCellExperiment)
library(SpatialExperiment)
library(OSTA.data)
library(SpatialExperimentIO)
library(scater)
library(BiocParallel)
```


## Example using `COMMOT`

Here, we demonstrate a short example using `COMMOT` [@Cang2023-COMMOT], an optimal transport-based approach that models competition between sender and receiver signals, considers physical distances, and easily links to public databases.

Since `COMMOT` is available as a Python package, we can access it from R using `r BiocStyle::Biocpkg("reticulate")` (see @sec-bkg-python-interoperability).

First, we set up the Python environment and install `COMMOT`:

```{r}
# set up Python environment using reticulate

# install Miniconda
#if (is.null(miniconda_path())) {
install_miniconda()
#}

env <- "img-ccc"
#if (!(env %in% conda_list()$name)) {
conda_create(env, packages = "python=3.11")
#}
use_condaenv(env, required = TRUE)

# install Python packages from conda-forge
# these are dependency packages for commot that need to be installed from conda-forge (not PyPI)
conda_install(
  envname = env, 
  packages = c("libcxx", "llvm-openmp", "llvmlite", "numba"), 
  channel = "conda-forge"
)

# install Python packages from PyPI
# commot is available from PyPI
conda_install(
  envname = env, 
  packages = c("commot==0.0.3", "anndata==0.12.2", "pandas==2.3.2"), 
  pip = TRUE
)

# run {python} chunks via reticulate
knitr::opts_chunk$set(python.reticulate = TRUE)
```

```{r}
# set up parallelization
bp <- MulticoreParam(4)
```


Load example dataset:

```{r load-data}
# load data
# retrieve dataset from OSF repository
id <- "CosMx6k_HumanBrain"
pa <- OSTA.data_load(id)
dir.create(td <- tempfile())
unzip(pa, exdir = td)
(spe <- readCosmxSXE(td, addTx = FALSE))
```

```{r preprocessing}
# preprocessing
spe <- logNormCounts(spe, BPPARAM = bp)

# wrangling
xy <- spatialCoords(spe)
reducedDim(spe, "spatial") <- xy
```


Run `COMMOT` using `reticulate` to run Python code:

```{r run-commot}
# run COMMOT using reticulate
ad <- import("anndata")
pd <- import("pandas")
ct <- import("commot")

# retrieve L-R interactions from CellChatDB and filter for those that are 
# fully covered by the 1k-plex CosMx panel
db <- ct$pp$ligand_receptor_database(database = "CellChat", species = "human")
names(db) <- c("ligand", "receptor", "pathway", "type")
nrow(db)

db <- db[apply(db, 1, \(.) {
  rs <- strsplit(.["receptor"], "_")
  lr <- c(.["ligand"], unlist(rs))
  all(lr %in% rownames(spe))
}), ]
nrow(db)

# subset to features being considered
rs <- sapply(strsplit(db$receptor, "_"), .subset, 1)
nrow(spe <- spe[unique(c(db$ligand, rs)), ])

is <- split(colnames(spe), spe$fov)
sr <- bplapply(is, BPPARAM = bp, \(.) {
  # skip FoV when there are too few cells
  if (length(.) < 100) return(NULL)
  ad <- SCE2AnnData(spe[, .], X_name = "logcounts")
  ct$tl$spatial_communication(
    ad, 
    database_name = "CellChatDB", 
    # assume average cell is 10x10um; here, we consider a 
    # distance threshold of 10 cells = 0.1mm
    dis_thr = 0.1, 
    df_ligrec = db, 
    heteromeric = TRUE, 
    pathway_sum = TRUE, 
    heteromeric_rule = "min", 
    heteromeric_delimiter = "_")
  list(
    ad$obsm["commot-CellChatDB-sum-sender"], 
    ad$obsm["commot-CellChatDB-sum-receiver"])
})
```


## Appendix

### References {.unnumbered}
