# Neighborhood analysis {#sec-img-neighborhood-analysis}

## Preamble

### Introduction

Neighborhood analysis aims to investigate the composition of and interaction between cells and cell types within small regions of tissue, which are referred to as neighborhoods or niches.


### Dependencies

```{r deps, message=FALSE, warning=FALSE}
library(dplyr)
library(tidyr)
library(scider)
library(ggplot2)
library(patchwork)
library(RColorBrewer)
library(BiocParallel)
library(SpatialExperiment)
```

```{r, message=FALSE, warning=FALSE}
# load data from previous section
spe <- readRDS("img-spe_cl.rds")
spe$k <- colLabels(spe)
```

```{r, message=FALSE, warning=FALSE}
# basic theme for spatial plots
theme_xy <- list(
    coord_equal(expand=FALSE), 
    theme_void(), theme(
        plot.margin=margin(l=5),
        legend.key=element_blank(),
        panel.background=element_rect(fill="black")))
```


## Nearest neighbors

Custom spatial analyses may rely on identifying nearest neighbors (NNs) of cells.
We recommend `r BiocStyle::CRANpkg("RANN")` for this purpose, which finds NNs in **$O(N\log N)$ time for $N$ cells** (by comparison, conventional approaches would take $O(N^2)$ time) by relying on a Approximate Near Neighbor (ANN) C++ library.
Furthermore, there is support for **exact, approximate, and fixed-radius searchers**.
The latter is of particular interest in biology; e.g., one might require $k$NNs to lie within a biologically sensible distance as to avoid consideration of cells that are far-off, especially in sparse regions or at tissue borders.

As a toy example, we here compute the $k$NNs between a pair of subpopulations, with and without thresholding on NN distances (`searchtype="radius"`).

- For the first approach, each cell will receive $k$ neighbors exactly,  
but these may lie within an arbitrary distance.

- For the second approach, cells will receive $\leq k$ neighbors,  
depending on how many cells lie within a radius $r$.

```{r nns-calc, message=FALSE, results="hold"}
library(RANN)
k <- 10 # num. neighbors
r <- 50 # dist. threshold
i <- spe$k == 1 # source
j <- spe$k == 4 # target
xy <- spatialCoords(spe)
# k-NN search: all cells have k neighbors
ns_k <- nn2(xy[j, ], xy[i, ], k=k)
is_k <- ns_k$nn.idx
all(rowSums(is_k > 0) == k) 
# w/ fixed-radius: cells have 0-k neighbors
ns_r <- nn2(xy[j, ], xy[i, ], k=k, searchtype="radius", r=r)
is_r <- ns_r$nn.idx
range(rowSums(is_r > 0))
```

The neighbors obtained via fixed-radius search (right) are less scattered than those obtained for unlimited distances (left); the former are arguably more meaningful in a biological context:

```{r nns-plot, fig.width=8, fig.height=3, fig.cap="Results of kNN searches. Left: basic kNN search, highlighting source (pink) and target cells (gold). Right: kNN search with same k, but thresholding on neighbor distances."}
#| code-fold: true
df <- data.frame(xy, colData(spe))
p0 <- ggplot(df, aes(x_centroid, y_centroid)) + 
    geom_point(data=df, col="navy", shape=16, size=0) +
    geom_point(data=df[i, ], col="magenta", shape=16, size=0) 
p1 <- p0 + geom_point(
    data=df[which(j)[is_k], ], 
    col="gold", shape=16, size=0) +
    ggtitle("k-nearest neighbors")
p2 <- p0 + geom_point(
    data=df[which(j)[is_r], ], 
    col="gold", shape=16, size=0) +
    ggtitle("fixed-radius search")
(p1 | p2) + plot_layout(nrow=1) & theme_xy
```

::: {.callout-note collapse="true" title="exhaustive fixed-radius search"}
Note that, we could also set a very large `k` in order to identify *all* neighbors within a radius `r`.
In order to prevent unnecessarily costly searches, it is sensible to estimate how many neighbors we would expect, and to set `k` accordingly. 
`nn2()` will otherwise find each cells' $k$NNs, and set the indices of those with a distance $>r$ to 0.

As an exemplary approach, we here sample 1,000 cells to estimate the highest number of NNs obtained, considering *half* of all target cells as potential NNs:

```{r nns-note-test}
# test search
.i <- sample(which(i), 1e3)
ns <- nn2(
    xy[j, ], xy[.i, ], 
    k=round(sum(j)/2), 
    searchtype="radius", r=r)
(.k <- max(rowSums(ns$nn.idx > 0)))
```

For our actual search, we then set `k` to be twice our estimate.
As a final spot-check, we make sure that all cells have fewer than `k` NNs, since we might otherwise be missing some.

```{r nns-note-real}
# real search
ns <- nn2(
    xy[j, ], xy[i, ], 
    k=k <- ceiling(2*.k), 
    searchtype="radius", r=r)
max(rowSums(ns$nn.idx > 0)) < k
```

:::


## Spatial contexts

Spatial niche analysis aims at identifying regions of homogeneous composition by grouping cells based on their microenvironment.
To this end, methods such as `r BiocStyle::Biocpkg("imcRtools")` [@Windhager2023-imcRtools] rely on a $k$-nearest-neighbor ($k$NN) graph (based on Euclidean cell-to-cell distances), and clustering cells using common clustering algorithms (according to their neighborhood's subpopulation frequencies).

Here, we demonstrate how to identify spatial contexts based on $k$-means clustering on cluster frequencies among (Euclidean) $k$NNs.
We recommend readers consult the `imcRtools` documentation for a much wider range of visualizations and downstream analyses in this context.

```{r imcRtools-calc, message=FALSE}
library(imcRtools)
# construct kNN-graph based on Euclidean distances
sqe <- buildSpatialGraph(spe, 
    coords=spatialCoordsNames(spe),
    img_id="sample_id", type="knn", k=10)
# compute cluster frequencies among each cell's kNNs
sqe <- aggregateNeighbors(sqe, 
    colPairName="knn_interaction_graph", 
    aggregate_by="metadata", count_by="k")
# view composition of 1st cell's kNNs
unlist(sqe$aggregatedNeighbors[1, ]) 
# cluster cells by neighborhood compositions
ctx <- kmeans(sqe$aggregatedNeighbors, centers=5)
table(sqe$ctx <- factor(ctx$cluster))
```

Let's quickly view the subpopulation composition of each spatial context:

```{r imcRtools-show}
df <- data.frame(spatialCoords(sqe), colData(sqe))
round(100*with(df, prop.table(table(k, ctx), 2)), 2)
```

Secondly, let's visualize the obtained spatial contexts in space:

```{r imcRtools-plot, fig.width=9, fig.height=3, fig.cap="Tissue plot with cells colored by cluster (left) and spatial context (right) based on $k$-means clustering of cluster frequencies among each cell's (Euclidean) $k$NNs."}
#| code-fold: true
pal_k <- unname(pals::trubetskoy(nlevels(df$k)))
pal_c <- c("blue", "cyan", "gold", "magenta", "maroon")
ggplot(df, aes(x_centroid, y_centroid, col=k)) + 
    scale_color_manual(values=pal_k) +
ggplot(df, aes(x_centroid, y_centroid, col=ctx)) + 
    scale_color_manual(values=pal_c) +
plot_layout(nrow=1) &
    geom_point(shape=16, size=0) &
    guides(col=guide_legend(override.aes=list(size=2))) &
    theme_xy & theme(legend.key.size=unit(0.5, "lines"))
```


## Co-localization

`r BiocStyle::Biocpkg("hoodscanR")` [@Liu2025-hoodscanR] also relies on a (Euclidean) $k$NN graph to estimate the probability of each cell associating with its NNs.
The resulting probability matrix (rows=cells, columns=NNs) can, in turn, be used to assess co-occurrence of subpopulations. 

```{r hoodscanR-calc, message=FALSE}
library(hoodscanR)
sqe <- readHoodData(spe, anno_col="k")
nbs <- findNearCells(sqe, k=100)
mtx <- scanHoods(nbs$distance)      
grp <- mergeByGroup(mtx, nbs$cells) 
sqe <- mergeHoodSpe(sqe, grp)       
```

To perform neighborhood co-localization analysis, `plotColocal()` computes the Pearson correlation of probability distribution between cells.
Here, **high/low values indicate attraction/repulsion** between clusters:

```{r hoodscanR-plot-corr, message=FALSE, fig.width=3.5, fig.height=3}
library(pheatmap)
cor <- plotColocal(sqe, pm_cols=colnames(grp), return_matrix=TRUE)
pal <- colorRampPalette(rev(hcl.colors(9, "Roma")))(100)
pheatmap(cor, 
    cellwidth=15, cellheight=15, 
    treeheight_row=5, treeheight_col=5,
    col=pal, breaks=seq(-1, 1, length=100))
```

::: {.callout-note title="measuring local mixing" collapse="true"}

Downstream, `calcMetrics()` can be used to calculate cell-level [entropy](https://en.wikipedia.org/wiki/entropy_(information_theory)) and [perplexity](https://en.wikipedia.org/wiki/perplexity),
which both measure the mixing of cellular neighborhoods.
Here, **low/high values indicate heterogeneity/homogeneity** of a cell's local neighborhood:

```{r hoodscanR-calc-mets}
sqe <- calcMetrics(sqe, pm_cols=colnames(grp))
```

```{r hoodscanR-plot-mets-xy, fig.width=8, fig.height=3, fig.cap="Tissue plots with cells colored by entropy and perplexity, z-scaled across cells (capped at 2 SDs)."}
#| code-fold: true
df <- data.frame(colData(sqe), k=spe$k, spatialCoords(spe))
vs <- c("perplexity", "entropy")
fd <- df |>
    mutate(across(all_of(vs), scale)) |>
    pivot_longer(all_of(vs)) |>
    # threshold at 2 SDs for clearer visualization
    mutate(value=case_when(abs(value) > 2 ~ sign(value)*2, TRUE ~ value))
ggplot(fd, aes(x_centroid, y_centroid, col=value)) +
    facet_grid(~name) + geom_point(shape=16, size=0) + 
    theme_xy + theme(legend.key.size=unit(0.5, "lines")) +
    scale_color_gradient2("z-scaled\nvalue", low="cyan", mid="navy", high="magenta") 
```

Stratifying these values by subpopulation, we can observe that clusters forming distinct aggregates in space are lowest in entropy/perplexity (i.e., the most homogeneous locally):

```{r hoodscanR-plot-mets-k, fig.width=7, fig.height=2.5, fig.cap="Boxplot of entropy and perplexity, stratified by cluster assignment and z-scaled across cells (capped at 2 SDs)."}
#| code-fold: true
ggplot(fd, aes(k, value, fill=k)) +
    facet_wrap(~name) + 
    scale_fill_manual(values=pal_k) + 
    geom_boxplot(outlier.stroke=0, key_glyph="point") +
    scale_y_continuous("z-scaled value", limits=c(-2, 2)) +
    guides(fill=guide_legend(override.aes=list(shape=21, size=2))) +
    theme_bw() + theme(
        axis.title.x=element_blank(),
        panel.grid.minor=element_blank(),
        legend.key.size=unit(0.5, "lines"))
```

:::

## Cellular density

The `r BiocStyle::Biocpkg("scider")` package provides functions to model 
the global density of cells across a tissue section, and allows for several 
downstream analyses, including:

- cellular co-localization analysis
- boundary detection analysis
- differential density analysis

All functions operate on `SpatialExperiment` objects, allowing seemless integration. 

Here, we will demonstrate how to:

i. estimate and visualize cellular density
i. stratify cells based on their density (via countours)
i. quantify how subpopulation composition changes with cellular density

```{r}
# non-valid names for clusters are problematic for 'scider',
# so that we use x1-16 instead of 1-16 as 'cell_type's
spe$cell_type <- paste0("x", spe$k)
```

```{r scider-est}
# perform kernel density estimation
# (here, using all clusters)
spe <- gridDensity(spe)
# the resulting 'DataFrame' contains 
# one 'density_x*' column per cluster
res <- metadata(spe)$grid_density
head(res[, seq_len(7)])
```

We can visualize cellular densities stratified by subpopulation, i.e.,
each panel panel depicts the density of one subpopulation:

```{r scider-, warning=FALSE, fig.width=6, fig.height=5}
#| code-fold: true
ps <- lapply(levels(spe$k), \(k) plotDensity(spe, k) + ggtitle(k))
wrap_plots(ps, nrow=4) & 
    theme_xy & theme(
        legend.position="none",
        plot.title=element_text(hjust=0.5))
```

`scider`'s `findROI()` can be used to detect regions of interest (ROIs) based
on the density of particular subpopulations. Specificelly, the method employs
graph-based community dection to cluster cells based on their density:

```{r scider-roi, fig.width=5, fig.height=4}
spe <- findROI(spe, coi="x1") # detect ROIs
res <- metadata(spe)$roi      # access results
nlevels(res$component)        # how many ROIs?
plotROI(spe) + coord_equal()  # visualize
```

After defining ROIs, we can test the relationship between any 
two subpopulations within each ROI, or overall but accounting 
for ROI variation. Results can be visualized in a heatmap where 

- rows/columns = subpopulations and 
- bin values = correlation of their cellular  
(low = avoidance, high = co-localization).

```{r scider-cor, fig.width=5, fig.height=4.5}
#| layout-ncol: 2
mod <- corDensity(spe)
plotCorHeatmap(mod, roi=10)
```

Next, we can divide cells into different levels of grid density:

```{r scider-contour, message=FALSE, fig.width=6, fig.height=4}
spe <- getContour(spe, coi="x1")
plotContour(spe, coi="x1") + coord_equal()
```

We can then annotate cells by their locations within each contour, and 
visualize their composition from lowest to highest cellular density;
for example, 

```{r scide-allocate, message=FALSE, fig.width=6, fig.height=4}
spe <- allocateCells(spe)
plotCellCompo(spe, coi="x1") + 
    coord_cartesian(expand=FALSE) +
    theme(legend.key.size=unit(0.8, "lines"))
```


## Appendix

### References {.unnumbered}
